---
title: "Exploratory Analysis of Spaceship Titanic"
author: "Monica Amezquita; Alejandro Hernandez; Hugo Marquez"
date: "2022-07-29"
output: pdf_document
---

## Load Relevant Libraries and Data

```{r echo=FALSE}
library(stringr)
library(corrplot)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Explore the Training Set

```{r}
ggpairs(train[-c(1,4,13)])  # ignore features with many levels (i.e. names)
```

```{r}
M <-cor(train)
corrplot(M, type="upper", order="hclust",
         col=brewer.pal(n=8, name="RdYlBu"))
```

# Percentage of Transported

```{r}
proportions <- table(train$Transported)/length(train$Transported)
percentages <- proportions*100

```

# Explore Expenditure vs. Transported and Expenditure vs. Crysleep colored by Transported

```{r}
ggplot(train, aes(x = Transported,y = Expenditure, fill = Transported)) + 
  geom_boxplot() +
  ylim(c(0,2000))

ggplot(train[!train$CryoSleep,], aes(x = CryoSleep,y = Expenditure, fill = Transported)) + 
  geom_boxplot() + 
  ylim(c(0,4000))
```

# Explore Count of Individuals in each Deck colored by Transported

```{r}
ggplot(train, aes(x = DeckNum, fill = Transported)) + 
  geom_bar() + 
  facet_wrap(vars(Transported))
```


# Explore Count of "Side" colored by Transported
```{r}
ggplot(train, aes(x = Side, fill = Transported)) + 
  geom_bar() + 
  facet_wrap(vars(Transported))
```

# Explore Histogran of Age colored by IsSoloTraveler
```{r}
ggplot(train, aes(x = Age, fill = IsSoloTraveler)) + 
  geom_histogram(binwidth = 1) + 
  facet_wrap(vars(Transported))
```

# Explore Histogram of Age colored by Transported
```{r}
ggplot(train, aes(x = Age, fill = Transported)) + 
  geom_histogram() +
  facet_wrap(vars(Transported))
```